---
title: "Exploring Energy, CO2 emmissions, and GDP growth"
author: "Your name goes here"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: true
    toc: yes
    toc_float: yes
---


```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```

```{r}
library(tidyverse)
library(wbstats)
library(skimr)
library(countrycode)
library(here)
library(scales)
library(wbstats)



library(plotly)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(patchwork)
library(mapview)
library(tmap)


library(showtext)
library(ggtext)
library(viridis)

# load fonts we will use
font_add_google("Montserrat", "Montserrat") 
font_add_google("Lato", "Lato")

## Automatically use showtext to render text for future devices
showtext_auto()

```

# Exploring energy,electricity, emissions, GDP
## Transition to net zero

Think about the questions below. What data do you need to answer them? Where can you find such data?

• How do countries generate electricity?

• How do countries differ in electrifiying their energy needs? Which countries have the highest % of primary energy being electricity? 

• What is the link between C02 emissions and GDP growth?

• Which countries have decoupled CO2 emissions from GDP growth?

• What is the link between electricity intensity and CO2 emissions?


## Exploring sources of electricity production, CO2 emissions, and GDP per capita.

There are many sources of data on how countries generate their electricity and their CO2 emissions. I would like you to create three graphs:

## 1. A stacked area chart that shows how your own country generated its electricity since 2000.

You will use

`geom_area(colour="grey90", alpha = 0.5, position = "fill")`

## 2. A scatter plot that looks at how CO2 per capita and GDP per capita are related

## 3. A scatter plot that looks at how electricity usage (kWh) per capita/day GDP per capita are related

We will get energy data from the Our World in Data website, and CO2 and GDP per capita emissions from the World Bank, using the `wbstats`package.

```{r}
#| message: false
#| warning: false


# Download co2 data
co2 <- "https://nyc3.digitaloceanspaces.com/owid-public/data/co2/owid-co2-data.csv" |> 
  read_csv() |> 
    filter(year >= 1990) 


# Download electricity data
energy <- "https://nyc3.digitaloceanspaces.com/owid-public/data/energy/owid-energy-data.csv" |> 
  read_csv()  |>  
  filter(year >= 1990) |> 
  drop_na(iso_code) |> 
  select(1:5,
         biofuel = biofuel_electricity,
         coal = coal_electricity,
         gas = gas_electricity,
         hydro = hydro_electricity,
         nuclear = nuclear_electricity,
         oil = oil_electricity,
         other_renewable = other_renewable_exc_biofuel_electricity,
         solar = solar_electricity,
         wind = wind_electricity, 
         electricity_demand,
         electricity_generation,
         electricity_share_energy,
         net_elec_imports,	# Net electricity imports, measured in terawatt-hours
         energy_per_capita,	# Primary energy consumption per capita, measured in kilowatt-hours	Calculated by Our World in Data based on BP Statistical Review of World Energy and EIA International Energy Data
         energy_per_gdp,	# Energy consumption per unit of GDP. This is measured in kilowatt-hours per 2011 international-$.
         per_capita_electricity #	Electricity generation per capita, measured in kilowatt-hours

  ) 

# Download data for GDP per capita  https://data.worldbank.org/indicator/NY.GDP.PCAP.PP.KD
# GDP per capita, PPP (constant 2021 international $)

gdp_percap <- wb_data(country = "countries_only", 
                      indicator = "NY.GDP.PCAP.PP.KD", 
                      start_date = 1990, 
                      end_date = 2023,
                      return_wide=FALSE) |> 
  filter(!is.na(value)) |> 
  #drop unwanted variables
  select(-c(unit, obs_status, footnote, last_updated)) |> 
  rename(year = date,
         GDPpercap = value)





```

Specific questions:

1. Produce a graph that shows % of electricity produced by differesources over the last few years
How would you turn `energy` to long, tidy format?

```{r}



energy_long <- energy |> 
  select(1:14) |> 
  pivot_longer(cols = 6:12,
               names_to = 'source',
               values_to = 'value') |> 
  drop_na(value) |> 
  drop_na(iso_code)
  


energy_long |> 
  filter(country == "Germany") |> 
  ggplot()+
  aes(x= year, y = value, fill = source)+
  geom_area(colour="grey90", alpha = 0.5, position = "fill")

energy_long |> 
  filter(country == "Germany") |> 
  mutate(source = fct_reorder(source, value, max)) |> 
  ggplot()+
  aes(x= year, y = value, fill = source)+
  geom_area(colour="grey90", alpha = 0.5, position = "fill")


```

2.  You may need to join these data frames
    -   Use `left_join` from `dplyr` to [join the tables](http://r4ds.had.co.nz/relational-data.html)
    -   To complete the merge, you need a unique *key* to match observations between the data frames. Country names may not be consistent among the three dataframes, so please use the 3-digit ISO code for each country
        -   An aside: There is a great package called [`countrycode`](https://github.com/vincentarelbundock/countrycode) that helps solve the problem of inconsistent country names (Is it UK? United Kingdom? Great Britain?). `countrycode()` takes as an input a country's name in a specific format and outputs it using whatever format you specify.

```{r}
country_names <- energy_long |> 
  distinct(iso_code, country) |> 
  select(iso_code, country)

join1 <- gdp_percap |> 

  left_join(country_names, by = c("iso3c" = "iso_code" )) |> 
  rename(country = country.y)


join2 <- join1  |>  
  left_join(energy, by = c("country", "year")) |> 
  select(-country.x) |> 
  left_join(co2)


library(countrycode)
join2 <- join2 |> 
  mutate(
    continent = countrycode(country, 
                            origin = 'country.name',
                            destination = "continent"),
    iso3c = countrycode(country, 
                        origin = 'country.name',
                        destination = "iso3c"),
  ) 

```

3.  Write a function that takes as input any country's name and returns all three graphs. You can use the `patchwork` package to arrange the three graphs

```{r}
library(viridis)
energy_mix <- function(country) {
  
  energy |> 
    select(1:2, 6:14) |> # we will now have 11 columns
    pivot_longer(cols = 3:11,
                 names_to = "source",
                 values_to = "value") |> 
    
    filter(country == {{country}}, 
           year >= 2000,
           value > 0 ) |> 
    
    mutate(source = fct_reorder(source, value, max)) |> 
    drop_na(value) |> 
    
    
    ggplot()+
    aes(x=year, y = value, fill=source )+
    geom_area(color="grey90", 
              alpha = 0.5,
              position = "fill")+ # so you get to 100%
    theme_bw()+
    scale_fill_viridis(discrete = T,
                       option = "H") +
    scale_y_continuous(labels = scales::percent)+
    labs(title = glue::glue("Electricity Production Mix for ", {{country}}),
         x=NULL,
         y=NULL,
         fill= "Source")
}

energy_mix("Germany")
energy_mix("France")
energy_mix("China")


```

A function that plots top n countries for a chosen fuel in a certain year

```{r}
  
energy_long <-   energy |> 
    select(1:2, 6:14) |> # we will now have 11 columns
    pivot_longer(cols = 3:11,
                 names_to = "source",
                 values_to = "value") 


top_fuel <- function(.source, .year, .n) {
  
  energy_long |> 
    
    
    filter(year == .year,
           source == .source) |>
    drop_na(value) |> 
    mutate(country  =  fct_reorder(country, value)) |>
    slice_max(order_by = value, n = .n) |>
    ggplot(aes(x = value, y = country)) +
    geom_col() +
    theme_minimal()+
    geom_text(
      aes(label = scales::number(value, accuracy = 0.1), x = value - .25),
      colour = "white",
      size = 4,
      hjust = 1
    ) +
    labs(
      title = paste0("Top ",.n," ", .source, " producing countries ", " in ",.year),
      subtitle = "Terra-Watthours (TWh)",
      x= NULL,
      y = NULL
    )+
    theme(
      plot.title.position = "plot",
  #    plot.title = element_textbox_simple(size=16),
      axis.title.y = element_text(angle = 0, vjust = 0.5,size=14),
      axis.text = element_text(size=12),
      legend.position = "none") +
    NULL
}

top_fuel("gas", 2018, 20)

top_fuel("coal", 2021, 10)

top_fuel("nuclear", 2021, 10)

top_fuel("hydro", 2020, 10)

top_fuel("wind", 2021, 10)

top_fuel("solar", 2023, 15)

```

## Three graphs

A function that takes a country as input and returns 3 graphs

```{r}
#| message: false
#| warning: false
library(viridis)
library(patchwork)
library(ggrepel)

energy_mix_co2_gdp <- function(country) {
  
  plot1 <- energy |> 
    select(1:2, 6:14) |> # we will now have 11 columns
    pivot_longer(cols = 3:11,
                 names_to = "source",
                 values_to = "value") |> 
    
    filter(country == {{country}}, year >= 2000) |> 
    drop_na(value) |> 
    mutate(source = fct_reorder(source, value, max)) |> 

    
    
    ggplot()+
    aes(x=year, y = value, fill=source )+
    geom_area(color="grey90", 
              alpha = 0.5,
              position = "fill")+ # so you get to 100%
    theme_bw()+
    scale_fill_viridis(discrete = T,
                       option = "H") +
    scale_y_continuous(labels = scales::percent)+
    labs(title = glue::glue("Electricity Production Mix for ", {{country}}),
         x=NULL,
         y=NULL,
         fill= "Source")
  
  plot2 <- join2 |> 
    select(year, country, GDPpercap, co2_per_capita, per_capita_electricity) |> 
    filter(country == {{country}}, year >= 2000) |> 
    
    ggplot()+
    aes(x=GDPpercap, y=co2_per_capita, label=year)+
    geom_point()+
    geom_text_repel(size=3)+
    theme_light() +
    scale_x_continuous(labels = scales::dollar)+
    labs(title = "CO2 vs GDP per capita",
         subtitle = "Greenhouse emissions in tonnes of CO2-equivalents per person",
         x = "GDP per capita",
         y = "CO2 emissions")
  
  plot3 <- join2 |> 
    select(year, country, GDPpercap, co2_per_capita, per_capita_electricity) |> 
    filter(country == {{country}}, year >= 2000) |> 
    
    ggplot()+
    aes(x=per_capita_electricity/365, y=co2_per_capita, label=year)+
    geom_point()+
    geom_text_repel(size=3)+
    theme_light()+
    labs(title = "CO2 emissions vs per capita electricity consumption",
         x = "Daily Electricity usage (kWh)",
         y = NULL)
  
  
# use patchwork to arrange plots
plot1 / (plot2 + plot3)
}

energy_mix_co2_gdp("India")

energy_mix_co2_gdp("United Kingdom")

energy_mix_co2_gdp("Brazil")

energy_mix_co2_gdp("France")

energy_mix_co2_gdp("China")

energy_mix_co2_gdp("Germany")

energy_mix_co2_gdp("France")

energy_mix_co2_gdp("Nigeria")

energy_mix_co2_gdp("Argentina")





```

## How much electricity does the average person use per day?"

```{r}
mycolours <- c("grey70","tomato")


join2  |> 
  drop_na(per_capita_electricity) |> 
  mutate(country = fct_reorder(country, per_capita_electricity),
         myfill = ifelse(country %in% c("China","Denmark","India", "United States", "Germany", "Nigeria"), TRUE, FALSE)) |> 
  filter(year == 2023, 
         population > 30e6) |> 
  ggplot()+
  aes(x=per_capita_electricity/365, 
      y = country, 
      fill = myfill,
      label = scales::number(per_capita_electricity/365,accuracy=0.1))+
  geom_col()+
  facet_wrap(~continent, scales="free")+
  scale_fill_manual(values = mycolours)+
  geom_text(
    colour = "#fafafa",
    size = 3,
    hjust = 1
  )+
  theme_minimal()+
  labs(
    title = "How much electricity does the average person use per day?",
    subtitle = "2023 kWh per capita, countries with population > 30m",
    fill = "Energy Source",
    x = NULL,
    y = NULL
  )+
  # ensure title is top-left aligned
  theme(plot.title.position = "plot")+
  theme(legend.position = "none")+
  NULL

```

```{r}
#| message: false
#| warning: false


# per capita kwh vs co2 per cap
co2_kwh_plot <- join2  |> 
  drop_na(per_capita_electricity) |> 
  filter(year == 2022, 
         population > 10e6) |> 
  ggplot()+
  aes(y=per_capita_electricity/365, x = co2_per_capita, colour = continent, 
        size = population^2, 
      label = country)+
  geom_point(alpha = 0.8)+
  labs(
    title = " CO2 per capita vs energy intensity",
    subtitle = "2022 kWh per capita, countries with population > 10m",
    x = "Greenhouse emissions per capita in CO2e",
    y = "kWh per capita per day"
  )+
  coord_flip()+
  theme_minimal(base_family ="Montserrat")+
  
  
  
  
  geom_text_repel(size = 4)+
  theme(legend.position = "none")+
  scale_colour_viridis(discrete = T,
                       option = "inferno")+
  theme(
    plot.title.position = "plot",
    # plot.title = element_textbox_simple(size=16),
    # axis.text = element_text(size=9),
    legend.position = "none") +
  NULL   


co2_kwh_plot

# Interactive CO2 vs kWh/capita plot with plotly

ggplotly(co2_kwh_plot)


```


## World map of CO2 emissions

```{r}
##  World map with CO2 per cap

map <- ne_countries(scale = "medium", returnclass = "sf") |>
  dplyr::select(name, iso_a3, geometry) |>
  filter(!name %in% c("Greenland", "Antarctica"))

df <- map |> 
  left_join(join2 |> 
              filter(year == 2021), 
            by = c("iso_a3" = "iso3c")) 



base_map <- ggplot(data = df) +
  geom_sf(
    mapping = aes(
      geometry = geometry, #use Natural Earth World boundaries
      fill = co2_per_capita #fill colour = percent vaccinated
    ),
    colour = "#FAFAFA",      # white borders between regions
  )+
  geom_sf_text(aes(label=scales::number(co2_per_capita, accuracy = 0.1)), size=2) + 
  scale_fill_gradientn(
    colours = c('#a6d96a','#d7191c'))+
  labs(title = "2021 CO2 emissions per capita")+
  theme_void()+
  theme(legend.position = "none")+
  NULL

base_map

# Mercator (ew)
map_mercator <- base_map +
  coord_sf(crs = "+proj=merc")

map_mercator


# Robinson projection
map_robinson <- base_map +
  coord_sf(crs = "+proj=robin") 

map_robinson

```


## Carbon Intensity of Electricity consumption

While all of the previous analysis looks at historical data over the last 30+ years, the [Electricity Maps website](https://app.electricitymaps.com/map) provides real-time data about carbon intensity around the world. What if one drives a Battery EV in Sweden or Poland? Both are good for the environment, right?

The electricity maps website makes its data avaiable too, and [Grant Chalmers create a great animation of carbon intensity of electricity consumption for Nov 2023](https://twitter.com/GrantChalmers/status/1728546415855702347)


# On your own

You have to come up with 2-3 plots using the data here plus any other data you may want to use to  about about electricity, energy, CO2 emissions, GDP Growth etc. 

A lot of graphs are given to you, the data has been cleaned, joined, etc.


> You have to come up with 2-3 plots using the data here plus any otehr data you may want to use to  about about electricity, energy, CO2 emissions, GDP Growth etc. 

Before you jump into making your plots, I would like you to think of the following

1.  **Audience (who)**

    -   To whom are you communicating?
    -   What do they already know?
    -   What is your position and relationship?

2.  **Content (what)**

    -   What do you want them to know or do?
    -   How will you communicate with them?
    -   What tone do you want your communication to set?

3.  **Evidence (how)**

    -   What data is available to make my point?

4.  **Context: Prepare Yourself**

    -   What is the one key massage they should take home?
    -   What background information is essential? What’s irrelevant?
    -   What are potential biases of (some of) the audience?
    -   What factors could weaken your case? Can we address them proactively?

A rubric for this assignment is posted on Canvas